Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-07-11 ltx/data_paper/DataPaper.tex
edits to disturbance support estimation
edits to result section

&\mathcal{F}^{-1}\left\lbrace \frac{1}{\tilde{m}(\boldsymbol\iota)}\left[\mathcal{F}\{\mathbf{E}[R_{y_{t+1}y_{t+1}}(\boldsymbol{\iota})]\}- \left[\frac{\mathcal{F}\{\mathbf{E}[R_{y_{t+1}y_t}(\boldsymbol{\iota})]\}\mathcal{F}\left\{\mathbf{E}\left[R_{y_ty_{t+1}}(\boldsymbol\iota)\right] \right\}}{\mathcal{F}\{(\mathbf{E}\left[R_{y_ty_t}(\boldsymbol\iota)\right]\} - \sigma_{\varepsilon}^2}\right]-\sigma_{\varepsilon}^2\right]\right\rbrace  ,

@@

From \eqref{eq:EM-MMGFourier2}, to calculate the exact shape of the disturbance covariance function, observation noise variance and the sensor kernel support are required. Again if the signal to noise ratio is high the effect of $\sigma_{\varepsilon}$ would be small. If $m$ is a point sensor then the spatial support of $\gamma$ can be calculated directly, when considering a sensor with a spatial extent, dividing by $\tilde{m}(\boldsymbol\iota)$ in \eqref{eq:EM-MMGFourier2} might generate numerical error.  However, assuming a Gaussian shape for disturbance function and sensor kernel (with known support), the support of the disturbance covariance function, $\sigma^2_d$ can be determined analytically using the expression for Gaussian basis functions' convolution, giving 

@@

This section demonstrates the identification of the proposed IDE model using simulated data. The observation noise and disturbance were unknown to the estimation algorithm and no assumption were made on the shape of the kernel.  Data was generated employing state space equation (\ref{eq:StateEq}-\ref{eq:QmatrixForSigmapoints}) using a square grid of $n_x= 9 \times 9$ equally spaced ($2.5~mm$) basis functions with $\sigma_{\phi}^2=1.58~mm^2$, assuming the observed field can be fully described by the decomposed IDE model. 100 realizations of 500 ms of data were generated and the first 100 ms was discarded allowing the model's dynamic to reach steady state. The observation noise was set to $\Sigma_{\epsilon}=0.2\times \mathbf I_{n_y}$ and the field disturbance was set to $\mathcal{GP}(\mathbf 0,0.1 \times\gamma(\mathbf{r}-\mathbf{r}'))$ where the covariance function was chosen as a Gaussian with zero mean and the width $\sigma_{\gamma}^2=1.7~mm^2$ . The distance between adjacent sensors was 1.5 mm resulting into $n_y = 196$ observations. The width of the sensor kernels were set to $\sigma_m^2=0.81~mm^2$. The sampling period and the synaptic time constant were set to $T_s = 1~ms$  and $\tau = 10~ms$, respectively. The sigmoidal activation parameters were $\varsigma=0.56~mV^{-1}$ and $v_0=1.8~mV$.

The correlation analysis described in Section \ref{subsec:CorrAnal} were applied to generated data in order to approximate the unknown supports of both connectivity kernel and the disturbance covariance function. In the algorithm, the unknown variables $\sigma_{\epsilon}^2$ and $\xi$ in equations () and () were set to zero. By doing so, the exact shape of the kernel and disturbance covariance function cannot be inferred, this step might be considered as a prerequisite for the EM algorithm, however, an approximate of their support can be still extracted. In fact, the estimated support for the disturbance covariance function was very accurate. The estimates of the connectivity kernel and disturbance support were normalised and thresholded for values less than 0.01. The result of the correlation analysis for the connectivity kernel is shown in Figure.\ref{fig:KernelWidthEstimation}. The red circle shows the estimated support which encompasses the non-zero values (black area) after thresholding. The estimated support of the disturbance covariance function convolved twice with the observation kernel is shown in Figure.\ref{fig:DisturbanceWidthEstimation}. By fitting a Guassian basis function to the result, and under the assumption that the width of the sensor,$\sigma_m^2$ were known, the width of the disturbance covariance function, $\sigma_d^2$ was calculated analytically using the expression for Gaussian basis functions convolution.

The estimated supports were then applied in the EM algorithm described in Section \ref{sec:EstimationAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the hyperparameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25~mm$ spacing were used to cover the estimated support where $\sigma_{\psi}^2=3.5$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The true and estimated connectivity kernel for 100 Monte Carlo simulations are shown in Figure~\ref{fig:KernelEstimates}. Cross-sections of the connectivity kernel estimates for each run are shown in Figure~\ref{fig:KernelEstimateCrossSection}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel is inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. Figure~\ref{fig:Histogram} depicts the histogram of parameter estimates from 100 simulations where the true parameter value is shown by the solid line and the mean parameter estimate is shown by the dashed line. These histograms indicate small bias in parameter estimates with a very small standard deviation. Examples of neural field reconstruction are given in Figure.~\ref{fig:FieldEstimates}.  The figure shows snapshots at three time instants of the true neural field and the estimated field. In each subplot neural activity is plotted against spatial index in 1D to provide a comparision between true and the estimated underlying fields. Plot of lower bound of the log likelihood versus iterations of the EM is shown in Figure.~\ref{fig:LowerBound}. In general, for the non-linear dynamical systems the actual lower bound can not be claculated analytically, therefore the approximate of the lower bound obtained by URTSS is shown. The change in lower bound drops below 0.009 percent after 15 iterations, confirming
that the algorithm converged to steady parameter values.
\section{Discussion}

>>>>

&\mathcal{F}^{-1}\left\lbrace \frac{1}{\tilde{m}(\boldsymbol\iota)}\left[\left[\mathcal{F}\{\mathbf{E}[R_{y_{t+1}y_{t+1}}(\boldsymbol{\iota})]\}-\sigma_{\varepsilon}^2\right]- \left[\frac{\mathcal{F}\{\mathbf{E}[R_{y_{t+1}y_t}(\boldsymbol{\iota})]\}\mathcal{F}\left\{\mathbf{E}\left[R_{y_ty_{t+1}}(\boldsymbol\iota)\right] \right\}}{\mathcal{F}\{(\mathbf{E}\left[R_{y_ty_t}(\boldsymbol\iota)\right]\} - \sigma_{\varepsilon}^2}\right]\right]\right\rbrace  ,

@@

From \eqref{eq:EM-MMGFourier2}, to calculate the exact shape of the disturbance covariance function, observation noise variance and the sensor kernel support are required. Again if the signal to noise ratio is high the effect of $\sigma_{\varepsilon}$ would be small. Note the first bracket and the denominator in \eqref{eq:EM-MMGFourier2} are non-negative power spectral denisties of the noise free observations at time $t+1$ and $t$ respectively. If $m$ is a point sensor then the spatial support of $\gamma$ can be calculated directly, when considering a sensor with a spatial extent, dividing by $\tilde{m}(\boldsymbol\iota)$ in \eqref{eq:EM-MMGFourier2} might generate numerical error.  However, assuming a Gaussian shape for disturbance function and sensor kernel (with known support), the support of the disturbance covariance function, $\sigma^2_d$ can be determined analytically using the expression for Gaussian basis functions' convolution, giving 

@@

This section demonstrates the identification of the proposed IDE model using simulated data. The observation noise and the field disturbance were unknown to the estimation algorithm and no assumption were made on the shape of the connectivity kernel. Data was generated employing state-space equation (\ref{eq:StateEq}-\ref{eq:QmatrixForSigmapoints}) using a square grid of $n_x= 9 \times 9$ equally spaced ($2.5~mm$) basis functions with $\sigma_{\phi}^2=1.58~mm^2$, assuming the observed field can be fully described by the decomposed IDE model. 100 realizations of 500 ms of data were generated and the first 100 ms was discarded allowing the model's dynamic to reach steady state. Observation noise variance was set to 0.2. All other parameters for the simulation were the same as Chapter 2 (see Table 1.1).

The correlation analysis described in Section \ref{subsec:CorrAnal} were applied to generated data in order to approximate the unknown supports of both connectivity kernel and the disturbance covariance function. However, to compute the supports the unknown parameters, $\sigma_{\epsilon}^2$ and $\xi$ are required and should be guessed beforehand.  To form a more accurate initial guess of the unkown parameters and explore the effect on the estimated support several tests were performed. First consider equation \eqref{eq:EM-MMGFourier2} for estimating the disturbance covariance function. As mentioned before the power spectral densities are non negative, therefore the minimum value of $\mathcal F \{\mathbf E\left[ R_{y_ty_t}\right] \}$ and $\mathcal F \{\mathbf E\left[ R_{y_{t+1}y_{t+1}}\right]\}$ should be chosen as the higher limit for the initial guess of $\sigma^2_{\varepsilon}$. The minimum value of the power spectral density was 0.23, Figure~\ref{fig:DisturbanceSupportExperiment} shows the support estimates for variyng  $\sigma^2_{\varepsilon}$ around the calculated limit. This figure shows that for values less than 0.23 the estimated support is close to the true one (see Figure~\ref{fig:BinaryDisturbanceWidthEstimation}(b)) while for values above 0.23 the shape of the disturbance covariance function is distorted. The estimated support of the covariance function convolved twice with the sensor kernel for  $\sigma^2_{\varepsilon}<$0.23 is shown in Figure~\ref{fig:BinaryDisturbanceWidthEstimation} (a), where the results were normalised and thresholded for values less than 0.01. By fitting a Guassian basis function to the result, and under the assumption that the width of the sensor, $\sigma_m^2$ were known, the width of the disturbance covariance function, $\sigma_d^2$ was calculated analytically using equation \eqref{eq:EM-AnalyticSupportCalculation}. Deconvolution can also be performed numerically to directly find the support as can be seen in Figure~\ref{fig:DisturbanceWidthEstimationCrossSection}.
%Figure 1 here
\begin{figure}[!ht]
\centering
\subfloat[][]{\includegraphics[scale=1]{./Figures/DisturbanceWidthEstimation0.pdf}}
\subfloat[][]{\includegraphics[scale=1]{./Figures/DisturbanceWidthEstimation01.pdf}}
\subfloat[][]{\includegraphics[scale=1]{./Figures/DisturbanceWidthEstimation02.pdf}}\\
\subfloat[][]{\includegraphics[scale=1]{./Figures/DisturbanceWidthEstimation023.pdf}}
\subfloat[][]{\includegraphics[scale=1]{./Figures/DisturbanceWidthEstimation03.pdf}}
\subfloat[][]{\includegraphics[scale=1]{./Figures/DisturbanceWidthEstimation04.pdf}}
\caption{{\bf Disturbance support estimates for variyng observation noise variance.} Each subplot shows the support of the disturbance covariance function convolved twice with the sensor kernel. (\textbf a) $\sigma_{\varepsilon}^2$=0. (\textbf b) $\sigma_{\varepsilon}^2$=0.1 . (\textbf c) $\sigma_{\varepsilon}^2$=0.2 .  (\textbf d) $\sigma_{\varepsilon}^2$=0.23 .  (\textbf e) $\sigma_{\varepsilon}^2$=0.3 . (\textbf f) $\sigma_{\varepsilon}^2$=0.4 .}
\label{fig:DisturbanceSupportExperiment}
\end{figure}
\begin{figure}[!ht]
\subfloat[][]{\includegraphics[scale=1]{./Figures/DisturbanceWidthEstimationBinary.pdf}}
\subfloat[][]{\includegraphics[scale=1]{./Figures/DisturbanceWidthTrue.pdf}}
\caption{{\bf Disturbance support approximation for }$\sigma_{\varepsilon}^2<$0.23. Each subplot shows the support of the disturbance covariance function convolved twice with the sensor kernel. (\textbf{a}) Estimated support is shown by red circle, nonzero values after thresholding are shown by the black area. (\textbf{b}) True support.}
\label{fig:BinaryDisturbanceWidthEstimation}
\end{figure}
\begin{figure}[!ht]
\centering
\includegraphics[scale=1]{./Figures/DisturbanceWidthEstimationCrossSection.pdf}
\caption{{\bf Cross-section of disturbance covariance support estimation}. The true and the estimated width are shown by solid and dashed lines respectively.}
\label{fig:DisturbanceWidthEstimationCrossSection}
\end{figure}

 By doing so, the exact shape of the kernel and disturbance covariance function cannot be inferred, this step might be considered as an initialisation for the EM algorithm, however, an approximate of their support can be still extracted. In fact, the estimated support for the disturbance covariance function was very accurate. The estimates of the connectivity kernel and disturbance support were normalised and thresholded for values less than 0.01. The result of the correlation analysis for the connectivity kernel is shown in Figure.\ref{fig:KernelWidthEstimation}. The red circle shows the estimated support which encompasses the non-zero values (black area) after thresholding. The estimated support of the disturbance covariance function convolved twice with the observation kernel is shown in Figure.\ref{fig:DisturbanceWidthEstimation}. By fitting a Guassian basis function to the result, and under the assumption that the width of the sensor, $\sigma_m^2$ were known, the width of the disturbance covariance function, $\sigma_d^2$ was calculated analytically using the expression for Gaussian basis functions convolution.

The estimated supports were then employed in the EM algorithm described in Section \ref{sec:EstimationAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the hyperparameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25~mm$ spacing were used to cover the estimated kernel support where $\sigma_{\psi}^2=3.5$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The true and estimated connectivity kernel for 100 Monte Carlo simulations are shown in Figure~\ref{fig:KernelEstimates}. Cross-sections of the connectivity kernel estimates for each run are shown in Figure~\ref{fig:KernelEstimateCrossSection}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel is inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. Figure~\ref{fig:Histogram} depicts the histogram of parameter estimates from 100 simulations where the true parameter value is shown by the solid line and the mean parameter estimate is shown by the dashed line. These histograms indicate small bias in parameter estimates with a very small standard deviation. Examples of neural field reconstruction are given in Figure.~\ref{fig:FieldEstimates}.  The figure shows snapshots at three time instants of the true neural field and the estimated field. In each subplot neural activity is plotted against spatial index in 1D to provide a comparision between true and the estimated underlying fields. Plot of lower bound of the log likelihood versus iterations of the EM is shown in Figure.~\ref{fig:LowerBound}. In general, for the non-linear dynamical systems the actual lower bound can not be claculated analytically, therefore the approximate of the lower bound obtained by URTSS is shown. The change in lower bound drops below 0.009 percent after 15 iterations, confirming
that the algorithm converged to steady parameter values.
\section{Discussion}
